---
title: "Regression Models - Automatic vs manual transmissions"
author: "chris-FR-GitHub"
date: "18 Décembre 2018"
output:
  html_document:
    keep_md: yes
  pdf_document: default
geometry: left=2cm,right=2cm,top=1cm,bottom=2cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set( fig.path = "figure/")
```

## Synopsis

In this week project, we will explore a data set containing a collection of cars and try to determine if there is a difference between automatic and manual transmissions in term of MPG. (The code of this document is available on Github)

## Data

```{r constants, message=FALSE, echo = FALSE}
library(tidyverse, warn.conflicts = TRUE)
library(gridExtra)
```

The data used for this project is the mtcars (Motor Trend Car Road Tests) dataset. It contains **`r dim(mtcars)[1]`** cars (rows) and  **`r dim(mtcars)[2]`** columns (**`r names(mtcars)`**).  
We will just convert AM and VS to factors.
```{r factor, message=FALSE}
mtcars$am <- factor(mtcars$am, levels = c(0, 1), labels = c("auto", "manual"))
mtcars$vs <- factor(mtcars$vs)
```

## Exploratory analysis

The **str** and **summary** function results are in the appendix. The repartition automatic and manual is the following:
```{r tableam, echo=FALSE}
cbind(count=table(mtcars$am), percentage=prop.table(table(mtcars$am))*100)
```
If we check the MPG value compared to the other column:  

```{r quickplot, echo=FALSE, fig.align='center', fig.height=3}
pam <- ggplot(mtcars, aes(x=am, y=mpg, fill=am)) + 
    geom_boxplot() + 
    theme(legend.position="none")

pcyl<- ggplot(mtcars, aes(x=factor(cyl), y=mpg, fill=factor(cyl))) + 
    geom_boxplot() + xlab('cyl') +
    geom_jitter(aes(col=mtcars$am)) +
    theme(legend.position="none")

pdisp<- ggplot(mtcars, aes(x=disp, y=mpg, col=am)) + 
    geom_point() + 
    theme(legend.position="none")

php<- ggplot(mtcars, aes(x=hp, y=mpg, col=am)) + 
    geom_point() + 
    theme(legend.position="none")

pdrat <- ggplot(mtcars, aes(x=drat, y=mpg, col=am)) + 
    geom_point() + 
    theme(legend.position="none")

pwt  <- ggplot(mtcars, aes(x=wt, y=mpg, col=am)) + 
    geom_point() + 
    theme(legend.position="none")

pqsec  <- ggplot(mtcars, aes(x=qsec, y=mpg, col=am)) + 
    geom_point() + 
    theme(legend.position="none")

pgear<- ggplot(mtcars, aes(x=factor(gear), y=mpg, fill=factor(gear))) + 
    geom_boxplot() + xlab('Gear') + 
    geom_jitter(aes(col=mtcars$am)) +
    theme(legend.position="none")

grid.arrange(pam, pcyl, pdisp, php, pdrat, pwt, pqsec, pgear, ncol=4, nrow = 2)
```

MPG is highly correlated to some features. The complete correlation plot is on appendix.
```{r correlations, echo=FALSE}
correlations <- cor(mtcars[,sapply(mtcars, class) == "numeric"], method="spearman")
round(correlations[1, -1],2)
```

## MPG vs AM

From the first graph, there seems to have a difference in MPG between the 2 transmissions. 
```{r ttest}
t<- t.test( mpg~am, data = mtcars, 
            alternative = "two.sided", paired = FALSE, var.equal = FALSE, conf.level = 0.95)
```
The p-value is **`r round(t$p.value, 4)`** and the 95% interval does not contains 0, so we can say that there is a difference in MPG between these 2 transmission types. if we create a model from this feature only :
```{r ammodel}
fitam <- lm(mpg~am, mtcars)
summary(fitam)$coefficient
```

Even if there is a difference between the 2 transmissions (manual having a **`r round(fitam$coefficients[2],2)`** higher MPG), this feature only is not a very good predictor : the Adjusted R-squared value is **`r round(summary(fitam)$adj.r.squared, 3)`** . Let's check if we can find a better model using the other features.

## Feature selection using step
In the feature selection video, the **step** function was indroduced. We will try to use it to find a better model and see if the transmission is part of it.I used : http://www.stat.columbia.edu/~martin/W2024/R10.pdf as a starter code.

```{r step}
fit.null<-lm(mpg~1, data=mtcars)
fit.full<-lm(mpg~., data=mtcars)
# trying the 3 ways
step_fw <- step(fit.null, scope=list(lower=fit.null, upper=fit.full), direction="forward", trace=0)
step_bc <- step(fit.full, data=mtcars, direction="backward", trace=0)
step_bo <- step(fit.null, scope = list(upper=fit.full), data=mtcars, direction="both", trace=0)
```
| direction | formula | Adjusted R-squared |
| ---- | ---- | ---- |
| forward | `r format(as.formula(step_fw))` | `r round(summary(step_fw)$adj.r.squared, 3)` |
| backward | `r format(as.formula(step_bc))` | `r round(summary(step_bc)$adj.r.squared, 3)` |
| both | `r format(as.formula(step_bo))` | `r round(summary(step_bo)$adj.r.squared, 3)` |

## Conclusion
If we pick the backward step result as our final model : **`r format(as.formula(step_bc))`**, Manual transmissions have a slightly better MPG (**`r round(coef(step_bc)[4], 2)`**) than the automatic ones (in the 1974's).
```{r final}
summary(step_bc)$coefficient
```
## Appendix

### Exploratory analysis

```{r strandsummary}
str(mtcars)
summary(mtcars[, sapply(mtcars, class) == "numeric"])
```

Pair plot:
```{r pair}
pairs(mpg~., data=mtcars, col=mtcars$am)
```
Correlation plot:
```{r corrplot, fig.height=3}
library(corrplot)
corrplot(correlations, method="circle")
```


