---
title: "Speed and Stopping Distances of Cars"
author: "Chris-FR"
date: "7 février 2019"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Overview

These slides were made using R Studio Presentation (ioslides). 

The Web application can be found at the URL:
https://chris-fr-github.shinyapps.io/DevDataPrd-Week4-ShinyWebApp/

Source code is at GitHub for the application and presentation at the URL:
https://github.com/chris-FR-GitHub/Coursera-Johns-Hopkins/tree/master/09-Developing-Data-Products/Week4

## Data
```{r lib, echo = FALSE, warning=FALSE,message=FALSE,comment=FALSE}
library(stats)
library(dplyr)
library(ggplot2)
library(plotly)
```
The data used for this application is `cars` from the `datasets` package.  
This dataset contains `r dim(cars)[1]` records of `r dim(cars)[2]` features.  
The data give the speed of cars and the distances taken to stop then (the data in from the 1920s).

```{r cars, echo = TRUE}
head(cars)
```

## Web application

The web application calculates linear / polynomial regression of the 2 features.  

The users can pick if he wants :  
- speed~dist or dist~speed  
- a linear regression or a polynomial regression (up to 5 degrees)  
- a value (the preticted value will be displayed).

The plot uses **plotly**.  
The numeric input field is dynamic and the max value depends of the formula.

## Exemple
Formula : dist~speed - 2 degree polynomial, value = 15
```{r plot, echo = FALSE, fig.height=3.5}
degree<-2
x<-cars$speed
y<-cars$dist
xlabel<- 'Speed'
ylabel<- 'Distance'
formula <- y ~ poly(x, degree, raw = TRUE)
p <- ggplot(cars, aes(x, y)) 
p <- p + 
    geom_point(alpha=0.2, 
                    fill="blue", colour="black", 
                    size=2) +
    labs(x = xlabel, y= ylabel)
p <- p + geom_smooth(method = "lm", 
                     se = FALSE, 
                     formula = formula, 
                     colour = "red")
ggplotly(p)
```

```{r predictedvalue, echo = FALSE}
lm.fit <- lm(formula = formula)
val <- predict(lm.fit, 
        newdata = data.frame(x = 15), se.fit = FALSE)
```
Predicted value for 15 : `r round(val,1)`
