---
title: "Statistical Inference - Project 1 - A simulation exercise"
author: "chris-FR-GitHub"
date: "21 novembre 2018"
output:
  pdf_document: default
  html_document:
    keep_md: yes
geometry: left=2cm,right=2cm,top=1cm,bottom=2cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set( fig.path = "figure/")
```

## Synopsis

In this first project, we will investigate the exponential distribution and compare it with the Central Limit Theorem.  
We will do a single sample, a montecarlo simulation and a bootstrap. For each of these section, we will compute the mean and the 95% confidence interval.  
We will use **lambda = 0.2** for all of the simulations.
The mean of exponential distribution is 1/lambda (= 5) and the standard deviation is also 1/lambda (= 5). We will investigate the distribution of averages of 40 exponentials and do a thousand simulations.  

```{r constants, message=FALSE}
library(tidyverse, warn.conflicts = TRUE)
# constants
lambda <- 0.2     # lambda for rexp
n <- 40           # sample size
nbsim <- 1000     # number of simulations
quantile = 0.975  # 95% with 2.5% on both sides of the range
# Seed
set.seed(1)    # set seed to replicate the results
```

## Single sample

We will pick a single sample, compute its mean and the 95% confidence interval.

```{r singlesample}
# pick single sample
singlesample <- rexp(n , lambda)

# compute the 95% confidence interval
mu <- mean(singlesample)
sigma <- sd(singlesample)
confidenceInterval = round(mu + c(-1, 1) * qt(quantile, df=n-1) * sigma / sqrt(n), 2)

# test : compute the same interval using the t.test function
# t.test(singlesample, conf.level = 0.95)$conf.int

paste('mean : ', round(mu, 2), 
      ' - 95% confidence interval : [', confidenceInterval[1], ' - ', confidenceInterval[2], ']')
```


## Montecarlo simulation

We will do 1000 simulations. For each of these simulations, we will store the mean and the 95% confidence interval. 
We add an extra column, indicating if the interval contains the expected mean : 5.

```{r mts}
mcs_means <- replicate(nbsim, { 
    dat <- rexp(n , lambda)
    tinterval <- t.test(dat, conf.level = 0.95)$conf.int
    c( mean = mean( dat ), int_inf = tinterval[1], int_max = tinterval[2],
       contains_exp_mean = tinterval[1] <= 5 & 5 <= tinterval[2])
}) 
mcs_means=as.tibble(t(mcs_means))
# add the row number as a column (for the graph)
mcs_means$idsim <- as.numeric(row.names(mcs_means))
head(mcs_means,3)

# we can check the number of samples containing the expected mean
table(mcs_means$contains_exp_mean)
```
942 sample confidence intervals contained the expected vale (5) => 94%. We can see this better by checking the first 100 simulation mean and confidence interval.
```{r mtserrorbargraph}
mcs_means %>% filter(idsim <= 100) %>% 
    ggplot(aes(x= idsim,y=mean, color=as.factor(contains_exp_mean))) +
    geom_point() +
    geom_hline(yintercept = 5, color="darkgreen", size=1) +
    geom_errorbar(aes(ymin=int_inf, ymax=int_max)) +
    labs(x = 'Simulation Id', y = 'Sample mean and 95% confidence interval',
        title = 'Sample means Hit - first 100 simulations' ) +
    theme(legend.position="none")

```

```{r mtsdistributiongraph}
binwidth = 0.25
ndmean <- mean(mcs_means$mean)
ndsd <- sd(mcs_means$mean)
ndvar <- var(mcs_means$mean)

mcs_means %>% ggplot(aes(x= mean)) + 
    geom_histogram(fill='lightblue', color='darkblue', binwidth = binwidth, size = 0.1) +
    geom_density(aes(y=binwidth * ..count..), alpha=.2, fill="red", color="red", size=.75) +
    theme_bw() +
    labs(x = 'Sample means', y = 'Frequency', title = 'Sample means' ) +
    geom_vline(xintercept = ndmean, color='darkgreen', size = 1) +
    stat_function(fun = function(x) dnorm(x, mean = ndmean, sd = ndsd) * nbsim * binwidth,
                  color = "darkgreen", size = 1)

```

As stated by the CLT, the distribution of the means (in red) is approximatively a standard normal (in green) of mean **`r round(ndmean,3)`** and sd **`r round(ndsd,3)`**.  
The variance **`r round(ndvar,3)`** is approximatively equal to  the variance of the distribution of the means of samples of 40 exponentials : (1/lambda^2/n) = **`r round(1/(lambda^2)/n,3)`**.

## Bootstrap (==just to test==)

We will just do a simple test of the bootstrap method. Pick a single sample, then resample it using the **boot** package.  

```{r bootstrap}
library(boot)
# pick a single sample
singlesample <- rexp(n , lambda)
# bootstrap
meanBoot <- boot(singlesample,
             function(x,i) mean(x[i]),
             R=1000)

# The mean based on the bootstrap method.
mean(meanBoot$t[,1])

# The variance based on the bootstrap method.
var(meanBoot$t[,1])

# compute the confidence interval with different method provided by the package
boot.ci(meanBoot,
        conf = 0.95,
        type = c("norm", "basic" ,"perc", "bca")
)
```

